\documentclass{article}
\usepackage{hyperref}
\begin{document}
\title{Exercise 3}
\author{Compilation 0368:3133}
\date{Due 6/12/2017}
\maketitle

\section{Introduction}
We continue our journey of building a compiler
for the invented object oriented language RioMare.
Remember that the entire specification of RioMare appears
inside the relevant folder of the course website.
In order to make this document self contained,
all the information needed to complete the third exercise is brought here again.

\section{Programming Assignment}
The third exercise implements a semantic analyzer that
scans the abstract syntax tree produced by
\href{http://www2.cs.tum.edu/projects/cup/}{CUP},
and checks if it contains any semantic errors.
The input for the semantic analyzer is a (single) text file containing a RioMare program,
and the output is a (single) text file indicating whether the input program
is semantically valid or not.
In addition to that, whenever the input program is valid semantically,
the semantic analyzer will add meta data to the abstract syntax tree,
which is needed for later phases (code generation and optimization).
The added meta data content will not be checked in exercises $3$,
but still, the best time to design and implement this addition is
exercise $3$.

\section{The RioMare Semantics}
To avoid unnecessary formalism, this section describes the semantics of RioMare
using a multitude of running examples. 
\subsection{Types}
The RioMare programming language defines two native types: integers and strings.
In addition, one can recursively define classes and arrays according to th
Binary operations are performed only between integers and strings.
A variable is a recursive entity, represented as a sub tree in the AST.
Assigning a value to a variable, 
Table \ref{Table_CFG_Of_RioMare} summarizes the context free grammar of RioMare.
You will need to feed this grammar to \href{http://www2.cs.tum.edu/projects/cup/}{CUP},
and make sure there are no shift-reduce conflicts.  

\begin{table}[h]
\centering
\begin{tabular}{ l c l }
  Program     & $::=$ & dec$^{+}$                                           \\ \\

  dec         & $::=$ & funcDec $|$ varDec $|$ classDec                       \\ \\

  varDec      & $::=$ & ID ID $[$ ASSIGN exp $]$ ';'                              \\
  funcDec     & $::=$ & ID ID $'('$ $[$ ID ID $[$ ',' ID ID $]^{*}$ $]$ $')'$
                              $'\{'$ stmt   $[$ stmt $]^{*}$ $'\}'$           \\
  classDec    & $::=$ & CLASS ID $[$ EXTENDS ID $]$
                               $'\{'$ cField $[$ cField $]^{*}$ $'\}'$      \\ \\

  exp         & $::=$ & $'('$ exp $')'$                                     \\
              & $::=$ & exp BINOP exp                                       \\
              & $::=$ & $[$ var '.' $]$ ID $'('$ $[$ exp $[$ ',' exp $]^{*}$ $]$ $')'$ \\
              & $::=$ & INT $|$ NIL $|$ STRING $|$ NEW ID $|$ var           \\ \\
  var         & $::=$ & ID                                                        \\
              & $::=$ & var '.' ID                                                \\
              & $::=$ & var $'['$ exp $']'$                                       \\ \\
  
  stmt        & $::=$ & varDec                                                             \\
              & $::=$ & var ASSIGN exp ';'                                                 \\
              & $::=$ & RETURN $[$ exp $]$ ';'                                             \\
              & $::=$ & IF $'('$ exp $')'$ $'\{'$ stmt $[$ stmt $]^{*}$ $'\}'$             \\
              & $::=$ & WHILE $'('$ exp $')'$ $'\{'$ stmt $[$ stmt $]^{*}$ $'\}'$          \\
              & $::=$ & $[$ var '.' $]$ ID $'('$ $[$ exp $[$ ',' exp $]^{*}$ $]$ $')'$ ';' \\ \\

  cField      & $::=$ & varDec $|$ funcDec                                             \\ \\
\end{tabular}
\caption{
Context free grammar for the RioMare programming language.
\label{Table_CFG_Of_RioMare}}
\end{table}

\begin{table}[h]
\centering
\begin{tabular}{ l l l l}
  int         & string  & void   & while \\
  class       & nil     & new    & if    \\
  PrintInt    & extends & return & end   \\
  PrintString &         &        &       \\
\end{tabular}
\caption{
Reserved keywords of RioMare.
\label{Table_Reserved_Keywords_of_RioMare}}
\end{table}

\begin{table}[h]
\centering
\begin{tabular}{|c|c|l|l| }
  \hline
  Precedence       & Operator & Description & Associativity \\
  \hline
  \hline
  1                & $:=$            & assign         & right \\
  \hline
  2                & $=$             & equals         & left  \\
  \hline
  3                & $<,\leq,>,\geq$ &                & left  \\
  \hline
  4                & $+,-$           &                & left  \\
  \hline
  5                & $*,/$           &                & left  \\
  \hline
  6                & $[$             & array indexing &       \\
  \hline
  7                & $($             & function call  &       \\
  \hline
  8                & $\verb"->"$     & field access   & left  \\
  \hline
\end{tabular}
\caption{
Binary operators of RioMare along with their associativity and precedence.
$1$ stands for the lowest precedence, and $9$ for the highest.
\label{Table_Binary_Operators_Of_RioMare}}
\end{table}

To create a graph visualization of the AST, please install graphviz
and run
\[
\verb"$ dot -Tjpeg -o ./AST_Graph.jpeg ./AST_Graph.txt"
\]
from \verb"EX5/LINUX_GCC_MAKE"

\section{Input}
The input for this exercise is a single text file, the input RioMare program.

\section{Output}
The output is a single text file that contains a tokenized representation of the input program.
Each token should appear in a separate line, together with the line number
it appeared on, and the character position inside that line.
The list of token names appears in Table 777,
and will only be used in this first exercise.
Later phases of the compiler will make no use of these token names.
\begin{table}[h]
\centering
\begin{tabular}{|l|c|c|l|c|}
  \hline
  Token Name & Description & & Token Name & Description \\
  \hline
  \hline
  LPAREN    & $($  & & ASSIGN           & $:=$                     \\
  RPAREN    & $)$  & & EQ               & $=$                      \\
  LBRACK    & $[$  & & LT               & $<$                      \\
  RBRACK    & $]$  & & GT               & $>$                      \\
  LBRACE    & $\{$ & & CLASS            &                          \\
  RBRACE    & $\}$ & & EXTENDS          &                          \\
  PLUS      & $+$  & & RETURN           &                          \\
  MINUS     & $-$  & & WHILE            &                          \\
  TIMES     & $*$  & & IF               &                          \\
  DIVIDE    & $/$  & &                  &                          \\
  COMMA     & $,$  & & INT$(value)$     & $value$ is an integer    \\
  DOT       & $.$  & & STRING$(value)$  & $value$ is a string      \\
  SEMICOLON & $;$  & & ID$(value)$      & $value$ is an identifier \\
  \hline
\end{tabular}
\caption{
Token names and printing format for the first exercise.
Each line in the output text file should contain a single token.
Note that three types of tokens are associated with corresponding values:
integers, identifiers and strings. The rest of the tokens encountered
should only contain their name,
the line number they appeared on, and the character position inside that line.
\label{Table_Token_Names_For_Exercise_1}}
\end{table}

Three types of tokens are associated with corresponding values: integers, identifiers and strings.
The printing format for these tokens can be easily deduced from the examples in Table \ref{Table_Token_Printing_Examples_For_Exercise_1}.
\begin{table}[h]
\centering
\begin{tabular}{|l|l| }
  Printed Lines Examples & Description \\
  \hline
  \hline
  INT($74$)[$3$,$8$] & integer $74$ is encountered in line $3$, character position $8$ \\
  \hline
  STRING(``Dan")[$2$,$5$] & string ``Dan" is encountered in line $2$, character position $5$ \\
  \hline
  ID(numPts)[$1$,$6$] & identifier numPts is encountered in line $1$, character position $6$ \\
  \hline
\end{tabular}
\caption{
Token names and printing format for the first exercise.
Each line in the output text file should contain a single token.
Note that three types of tokens are associated with corresponding values:
integers, identifiers and strings. The rest of the tokens encountered
should only contain their name,
the line number they appeared on, and the character position inside that line.
\label{Table_Token_Printing_Examples_For_Exercise_1}}
\end{table}

\section{Submission Guidelines}
The code for this exercise resides as usual in subdirectory EX5 of the course GitHub.
Currently, the grammar in RioMare.y contains a shift/reduce conflict,
so you should start by fixing this.
Next, you need to add the relevant derivation rules and AST constructors
for classes. Last, you should implement the missing parts of the RioMare semantic analyzer.
The semantic analyzer resides in the file semant.c, and this is where
most of your changes will occur.
Please submit your exercise in your GitHub repository under COMPILATION/EX5,
and have a makefile there to build a runnable program called compiler.
Make sure that compiler is created in the same level as the makefile: inside EX5.
To avoid the pollution of EX5, please remove all *.o files once the target is built.
The next paragraph describes the execution of compiler.

\paragraph{Execution parameters}
compiler receives $2$ input file names:\\ \\
InputRioMareProgram.txt\\
OutputStatus.txt

\end{document}
