\documentclass{article}
\usepackage{hyperref}
\begin{document}
\title{Exercise 2}
\author{Compilation 0368:3133}
\date{Due 30/10/2017}
\maketitle

\section{Introduction}
We continue our journey of building a compiler
for the invented object oriented language RioMare.
Remember that the entire specification of RioMare appears
inside the relevant folder of the course website.
In order to make this document self contained,
all the information needed to complete the second exercise is brought here again.

\section{Programming Assignment}
The second exercise implements a \href{http://www2.cs.tum.edu/projects/cup/}{CUP} based
parser on top of your \href{http://jflex.de/}{JFlex} scanner from the exercise $1$.
The input for the parser is a single text file containing a RioMare program,
and the output is a (single) text file indicating whether the input program
is valid syntactically or not.
The course repository contains a simple skeleton program,
and you are encouraged to work your way up from there.

\section{Lexical Considerations}
\begin{table}[h]
\centering
\begin{tabular}{ l c l }
  Program     & $::=$ & dec$^{+}$                                           \\ \\

  dec         & $::=$ & funcDec $|$ varDec $|$ classDec                       \\ \\

  varDec      & $::=$ & ID ID $[$ ASSIGN exp $]$ ';'                              \\
  funcDec     & $::=$ & ID ID $'('$ $[$ ID ID $[$ ',' ID ID $]^{*}$ $]$ $')'$
                              $'\{'$ stmt   $[$ stmt $]^{*}$ $'\}'$           \\
  classDec    & $::=$ & CLASS ID $[$ EXTENDS ID $]$
                               $'\{'$ cField $[$ cField $]^{*}$ $'\}'$      \\ \\

  exp         & $::=$ & $'('$ exp $')'$                                     \\
              & $::=$ & exp BINOP exp                                       \\
              & $::=$ & $[$ var '.' $]$ ID $'('$ $[$ exp $[$ ',' exp $]^{*}$ $]$ $')'$ \\
              & $::=$ & INT $|$ NIL $|$ STRING $|$ NEW ID $|$ var           \\ \\
  var         & $::=$ & ID                                                        \\
              & $::=$ & var '.' ID                                                \\
              & $::=$ & var $'['$ exp $']'$                                       \\ \\
  
  stmt        & $::=$ & varDec                                                             \\
              & $::=$ & var ASSIGN exp ';'                                                 \\
              & $::=$ & RETURN $[$ exp $]$ ';'                                             \\
              & $::=$ & IF $'('$ exp $')'$ $'\{'$ stmt $[$ stmt $]^{*}$ $'\}'$             \\
              & $::=$ & WHILE $'('$ exp $')'$ $'\{'$ stmt $[$ stmt $]^{*}$ $'\}'$          \\
              & $::=$ & $[$ var '.' $]$ ID $'('$ $[$ exp $[$ ',' exp $]^{*}$ $]$ $')'$ ';' \\ \\

  cField      & $::=$ & varDec $|$ funcDec                                             \\ \\
\end{tabular}
\caption{
Context free grammar for the RioMare programming language.
\label{Table_CFG_Of_RioMare}}
\end{table}

\paragraph{Identifiers} may contain letters and digits, and must start with a letter.
\paragraph{The keywords} in Table \ref{Table_Reserved_Keywords_of_RioMare} can not
be used as identifiers.
\begin{table}[h]
\centering
\begin{tabular}{ l l l l}
  int         & string  & void   & while \\
  class       & nil     & new    & if    \\
  PrintInt    & extends & return & end   \\
  PrintString &         &        &       \\
\end{tabular}
\caption{
Reserved keywords of RioMare.
\label{Table_Reserved_Keywords_of_RioMare}}
\end{table}
\paragraph{White spaces} consist of spaces, tabs and newlines
characters. They may appear between any tokens.
Keywords and identifiers must be separated by a white space,
or a token that is neither a keyword nor an identifier.
\paragraph{Comments} in RioMare are similar to those used in the C programming language:
a comment beginning with the characters // indicates that the remainder of the line is a comment.
In addition, a comment can be a sequence of characters that begins with /*,
followed by any characters, including newlines, up to the first occurrence of the end sequence */. Unclosed comments are lexical errors.
\paragraph{Integer} literals may start with an optional negation sign $-$,
followed by a sequence of digits.
Non-zero numbers should \textit{not} have leading zeroes.
Though integers are stored as 32 bits in memory,
they are artificially limited in RioMare to have
16-bits signed values between $-2^{15}$ and $2^{15}-1$.
Integers out of this range are lexical errors.

\begin{table}[h]
\centering
\begin{tabular}{|c|c|l|l| }
  \hline
  Precedence       & Operator & Description & Associativity \\
  \hline
  \hline
  1                & $:=$            & assign         & right \\
  \hline
  2                & $=$             & equals         & left  \\
  \hline
  3                & $<,\leq,>,\geq$ &                & left  \\
  \hline
  4                & $+,-$           &                & left  \\
  \hline
  5                & $*,/$           &                & left  \\
  \hline
  6                & $[$             & array indexing &       \\
  \hline
  7                & $($             & function call  &       \\
  \hline
  8                & $\verb"->"$     & field access   & left  \\
  \hline
\end{tabular}
\caption{
Binary operators of RioMare along with their associativity and precedence.
$1$ stands for the lowest precedence, and $9$ for the highest.
\label{Table_Binary_Operators_Of_RioMare}}
\end{table}

To create a graph visualization of the AST, please install graphviz
and run
\[
\verb"$ dot -Tjpeg -o ./AST_Graph.jpeg ./AST_Graph.txt"
\]
from \verb"EX5/LINUX_GCC_MAKE"

\section{Input}
The input for this exercise is a single text file, the input RioMare program.

\section{Output}
The output is a single text file that contains a tokenized representation of the input program.
Each token should appear in a separate line, together with the line number
it appeared on, and the character position inside that line.
The list of token names appears in Table 777,
and will only be used in this first exercise.
Later phases of the compiler will make no use of these token names.
\begin{table}[h]
\centering
\begin{tabular}{|l|c|c|l|c|}
  \hline
  Token Name & Description & & Token Name & Description \\
  \hline
  \hline
  LPAREN    & $($  & & ASSIGN           & $:=$                     \\
  RPAREN    & $)$  & & EQ               & $=$                      \\
  LBRACK    & $[$  & & LT               & $<$                      \\
  RBRACK    & $]$  & & GT               & $>$                      \\
  LBRACE    & $\{$ & & CLASS            &                          \\
  RBRACE    & $\}$ & & EXTENDS          &                          \\
  PLUS      & $+$  & & RETURN           &                          \\
  MINUS     & $-$  & & WHILE            &                          \\
  TIMES     & $*$  & & IF               &                          \\
  DIVIDE    & $/$  & &                  &                          \\
  COMMA     & $,$  & & INT$(value)$     & $value$ is an integer    \\
  DOT       & $.$  & & STRING$(value)$  & $value$ is a string      \\
  SEMICOLON & $;$  & & ID$(value)$      & $value$ is an identifier \\
  \hline
\end{tabular}
\caption{
Token names and printing format for the first exercise.
Each line in the output text file should contain a single token.
Note that three types of tokens are associated with corresponding values:
integers, identifiers and strings. The rest of the tokens encountered
should only contain their name,
the line number they appeared on, and the character position inside that line.
\label{Table_Token_Names_For_Exercise_1}}
\end{table}

Three types of tokens are associated with corresponding values: integers, identifiers and strings.
The printing format for these tokens can be easily deduced from the examples in Table \ref{Table_Token_Printing_Examples_For_Exercise_1}.
\begin{table}[h]
\centering
\begin{tabular}{|l|l| }
  Printed Lines Examples & Description \\
  \hline
  \hline
  INT($74$)[$3$,$8$] & integer $74$ is encountered in line $3$, character position $8$ \\
  \hline
  STRING(``Dan")[$2$,$5$] & string ``Dan" is encountered in line $2$, character position $5$ \\
  \hline
  ID(numPts)[$1$,$6$] & identifier numPts is encountered in line $1$, character position $6$ \\
  \hline
\end{tabular}
\caption{
Token names and printing format for the first exercise.
Each line in the output text file should contain a single token.
Note that three types of tokens are associated with corresponding values:
integers, identifiers and strings. The rest of the tokens encountered
should only contain their name,
the line number they appeared on, and the character position inside that line.
\label{Table_Token_Printing_Examples_For_Exercise_1}}
\end{table}

\section{Submission Guidelines}
The code for this exercise resides as usual in subdirectory EX5 of the course GitHub.
Currently, the grammar in RioMare.y contains a shift/reduce conflict,
so you should start by fixing this.
Next, you need to add the relevant derivation rules and AST constructors
for classes. Last, you should implement the missing parts of the RioMare semantic analyzer.
The semantic analyzer resides in the file semant.c, and this is where
most of your changes will occur.
Please submit your exercise in your GitHub repository under COMPILATION/EX5,
and have a makefile there to build a runnable program called compiler.
Make sure that compiler is created in the same level as the makefile: inside EX5.
To avoid the pollution of EX5, please remove all *.o files once the target is built.
The next paragraph describes the execution of compiler.

\paragraph{Execution parameters}
compiler receives $2$ input file names:\\ \\
InputRioMareProgram.txt\\
OutputStatus.txt

\end{document}
